Pattern matching for\\ $(213,231)$-avoiding permutations

### 2015-02-09 article.tex
modifcation : take in account of corner in bivincular

\begin{abstract}
	Given permutations $\sigma \in S_k$ and $\pi \in S_n$, the
  \emph{pattern containment} problem is to decide whether $\pi$ contains
  $\sigma$ as an order-isomorphic subsequence.
	We give a linear-time algorithm in case both $\pi$ and $\sigma$ avoid
	$213$ and $231$.
	For the special where only $\sigma$ avoids $213$ and $231$, we present a
	$O(kn^2)$ time algorithm. We extend our research to $(213,231)$-avoiding
	bivincular pattern and present a $O(kn^3)$ time algorithm.
\end{abstract}

@@

	A permutation $\pi$ is said to contain another permutation $\sigma$,
	in symbols $\sigma \preceq \pi$,
	if there exists a subsequence of entries of $\pi$ that has the same relative
	order as $\sigma$, and in this case $\sigma$ is said to be a
	\emph{pattern} of $\pi$.
	Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
	For example a permutation contains the pattern $123$ (resp. $321$) if it has
	an increasing (resp. decreasing) subsequence of length $3$.
	Here, note that members need not actually to be consecutive,
	merely increasing (resp. descreasing).
	As another example,
	$6152347$ contains $213$ but not $231$.
	During the last decade, the study of the pattern containment on permutations has
	become a very active area of research and
	a whole annual conference (\textsc{Permutation Patterns}) is now devoted
	to this topic.

@@

	A few particular cases of the pattern containment problem have been attacked successfully.
	The case of increasing patterns is solvable in
	$O(|\pi| \log \log |\sigma|)$ time in the RAM model \cite{Crochemore:Porat:2010},
	improving the previous 30-year bound of $O(|\pi| \log |\sigma|)$.
	(The algorithm also improves on the previous
	$O(|\pi| \log \log |\pi|)$ bound.)
	Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear-time
	by stack sorting algorithms.
	Any pattern of length $4$ can be detected in $O(|\pi| \log |\pi|)$ time
	\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
	Algorithmic issues for $321$-avoiding patterns containment has been investigated in
	\cite{Guillemot:Vialette:ISAAC:2009}.
	The pattern containment problem is also solvable in
	polynomial-time for separable patterns \cite{Ibarra:1997,Bose:Buss:Lubiw:1998}
	(see also \cite{Bouvel:Rossin:Vialette:CPM:2007} for LCS-like issues
	of separable permutations).
	Separable permutations are those permutations that contain neither
	$2413$ nor $3142$, and they are enumerated by the Schr{\"o}der numbers
	(sequence A006318 in OEIS).
	To the best of our knowledge,
	separable permutations first arose in the work of
	Avis and Newborn~\cite{Avis:Newborn:1981},
	who showed that they are precisely the permutations which can be sorted by an
	arbitrary number of pop-stacks in series,
	where a pop-stack is a restricted form of stack in which any pop operation
	pops all items at once.
	(Notice that the separable permutations include as a special case the
	stack-sortable permutations, which avoid the pattern $231$.)

@@

A \emph{permutation} of length $n$ is a one-to-one function from an
$n$-element set to itself.
We write permutations as words
$\pi = \pi_1\pi_2\,\ldots\,\pi_n$, whose letters are distinct
and uasually consist of the integers $12\,\ldots\,n$.
We let
$\pi[i]$ stands for $\pi_i$.
For the sake of convenience, we let
$\pi[i:j]$ stand for
$\pi_i\pi_{i+1}\,\ldots\,\pi_j$,
$\pi[:j]$ stand for $\pi[1:j]$ and
$\pi[i:]$ stand for $\pi[i:n]$.
As usual, we let $S_n$ denote the set of all permutations of length $n$.

@@

A \emph{bivincular pattern} (abbreviated BVP) $\sigma$
of length $k$ is a permutation in $S_k$ written in
two-line notation
(that is the top row is $12\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\sigma_2\,\ldots\,\sigma_k$).
We have the following conditions on the top and bottom rows
of $\sigma$:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the letters corresponding to
	$\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j$ in an occurrence of
	$\sigma$ in a permutaton must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive letters.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$
	in a permutation $\pi$ must begin with the leftmost
	letter of $\pi$,
	and
	if the bottom row of $\sigma$ begins with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$
	in a permutation $\pi$ must end with the rightmost
	letter of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the letters corresponding to
	$\sigma_i, \sigma_{i+1}, \ldots, \sigma_j$ in an
	occurrence of $\sigma$ in a permutation must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined
	letters.
	Moreover, if the top row of $\sigma$ begins with
	$^\ulcorner{1}$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must begin with
	the smallest letter of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must end with
	the largest letter of $\pi$.
\end{itemize}
Given a motif $\pmotif$ we represent the elements underlined by the set $\x$. An index $i$ of an element is in $\x$ if and only if $i$ and $i+1$ need to be adjacent ($\pmotif[i]$ and $\pmotif[i+1]$ are underlined).
For example if we have $\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$, then ${i,i+1,\ldots,j-1} \in \y$.
We represent the index that need to be overlined by the set $\y$. An index $i$ is in $\y$ if and only if $i$ and $i+1$ need to be adjacent ($i$ and $i+1$ are overlined).
For example if we have
For example if we have $\overline{i\,i+1\,\ldots\,j}$, then ${i,i+1,\ldots,j-1} \in \y$.
% A bivincular pattern avoids a set of permutations $P$ if its bottom
% row $\sigma_1\sigma_2\,\ldots\,\sigma$ avoids every permutation in $P$.

@@

This section focus on the pattern containment problem
in case only the pattern $\sigma$ avoids $231$ and $213$.

\begin{proposition}
	\label{Proposition:sigma avoids 213 and 231}
	Let $\sigma \in \AV_k(213,231)$ and $\pi \in S_n$.
	One can decide in $O(kn^2)$ time
	and $O(kn^2)$ space if $\pi$ contains $\sigma$.
\end{proposition}

For the proof and the algorithm we introduce a decomposition : we decompose the stripe of the motif into factor of same letter, and label those factor from right to left. We also need to add the last index into the first factor. For each factor, we denote the left most element by $\firsta$. For example, given the motif $981237654$, $\factor{1} =7654$, $\first{1}=7$, $\factor{2}=123$, $\first{2}=1$, $\factor{3}=98$ and $\first{9}$.

\begin{proof}[of Proposition~\ref{Proposition:sigma avoids 213 and 231}]
Given an ascent (descent) factor $\factor{i}$ of a motif $\pmotif$, if we want to match the suffix of $\pmotif$ starting by $\factor{i}$, $\factor{i}\factor{i-1}\cdots\factor{1}$ and given that we know every match of $\factor{i-1}\cdots\factor{1}$, the optimal match is to choose the match of $\factor{i-1}\cdots\factor{1}$ that minimize (maximize) the maximal (minimal) element of the match of $\factor{i-1}\cdots\factor{1}$ and that allow a match of factor $\factor{i}$. In other word if we want to match $\pmotif[\first{i}:]$ with $\ptext[j:]$ such that $\first{i}$ is matched with element at index $j$ in $\ptext$, the optimal way of matching $\pmotif[\first{i-1}:]$ is to choose the minimal (maximal) element at index $j'$ of $\ptext[j:]$ such that : (1) $\pmotif[\first{i-1}:]$ is matched with $\ptext[j':]$ and element $\first{i-1}$ is matched with element at index $j'$, (2) the factor $\factor{i}$ is matched with $\ptext[j,j'-1]$ and element $\first{i}$ is matched with element at index $j$, (3) Every elements of the match of (2) if inferior (superior) to the minimal (maximal) element of the match of (1). Indeed when we will match $\factor{i+1}\factor{i}\factor{i-1}\cdots\factor{1}$, every element of $\factor{i+1}$ must be matched to element superior (inferior) to every element matched element of $\factor{i}\factor{i-1}\cdots\factor{1}$, particularly its must be superior (inferior) to the maximal (minimal) element of the match of $\factor{i}\factor{i-1}\cdots\factor{1}$.

@@

There exists a match of $\pmotif$ in $\ptext$ if and only if
there exists a $LM(n,i)\neq 0$ for $1 \leq i \leq |\ptext|$,
with $n$ the number of factor in $\pmotif$.
Moreover the basic case can be computed in $O(|\ptext|^2)$
and the induction on $O(|\pmotif||\ptext|^2)$.
\qed
\end{proof}

@@

\section{$(213,231)$-avoiding bivincular patterns}
	\label{section:bivincular}

		This section is devoted to presenting a polynomial-time algorithm for deciding whether
	a $(213,231)$-avoiding bivincular pattern occurs in a permutation.

	\begin{proposition}
		\label{Proposition:bivincular pattern}
			Let $\sigma$ be a $(213,231)$-avoiding bivincular pattern of length $k$
			and $\pi \in S_n$.
			One can decide in $O(kn^3)$ time
			and $O(kn^3)$ space if $\pi$ contains $\sigma$.
	\end{proposition}

@@

$
\PM{\pbmotif}{\ptext}{\min}{\max}{i}{j}=
\begin{cases}
	true 	& \text{If $\pbmotif[i:]$ is a motif of $\ptext[j:]$}\\
			& \text{with every element in}\\
			&[\min,\max]\\
>>>>>>> f9bc0a20a19c155782e9e48e9d01a9fb6f3718ac

@@

There are two cases to consider searching for a match :
\begin{itemize}
	\item If $n_\pmotif \in \x$ there exists a match if and only if there is a match starting with the first element of the text:\\ $\bigcup_{0<\min<\max<n_\ptext}\mathbb{P}_{\pmotif,\ptext,\min,\max}(0,0)$ is true.
	\item If $n_\pmotif \notin \x$ there exists a match if and only if there exists a match stating at any index: \\	$\bigcup_{0<k<n_\ptext,0<\min<\max<n_\ptext}\mathbb{P}_{\pmotif,\ptext,\min,\max}(0,k)$ is true.
\end{itemize}

\end{proof}
>>>>>>> f9bc0a20a19c155782e9e48e9d01a9fb6f3718ac

>>>>

\begin{abstract}
	Given permutations $\sigma \in S_k$ and $\pi \in S_n$, the Permutation Pattern (PP)
	problem is to decide whether $\pi$ contains $\sigma$ as an order-isomorphic pattern.
	We give a linear-time algorithm in case both $\pi$ and $\sigma$ avoid 
	$213$ and $231$. 
	For the special where only $\sigma$ avoids $213$ and $231$, we present a 
	$O(kn^2)$ time algorithm. We extend our research to $(213,231)$-avoiding 
	bivincular pattern and present a $O(kn^3)$ time algorithm.
\end{abstract}

@@

	A permutation $\pi$ is said to contain another permutation $\sigma$,
	in symbols $\sigma \preceq \pi$,
	if there exists a subsequence of entries of $\pi$ that has the same relative
	order as $\sigma$, and in this case $\sigma$ is said to be a 
	\emph{pattern} of $\pi$.
	Otherwise, $\pi$ is said to avoid the permutation $\sigma$.
	For example a permutation contains the pattern $123$ (resp. $321$) if it has
	an increasing (resp. decreasing) subsequence of length $3$.
	Here, note that members need not actually to be consecutive, 
	merely increasing (resp. descreasing).
	As another example,
	$6152347$ contains $213$ but not $231$.
	During the last decade, the study of the pattern containment on permutations has
	become a very active area of research and
	a whole annual conference (\textsc{Permutation Patterns}) is now devoted 
	to this topic.

@@

	A few particular cases of the pattern containment problem have been attacked successfully.
	The case of increasing patterns is solvable in
	$O(|\pi| \log \log |\sigma|)$ time in the RAM model \cite{Crochemore:Porat:2010},
	improving the previous 30-year bound of $O(|\pi| \log |\sigma|)$.
	(The algorithm also improves on the previous
	$O(|\pi| \log \log |\pi|)$ bound.)
	Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear-time
	by stack sorting algorithms.
	Any pattern of length $4$ can be detected in $O(|\pi| \log |\pi|)$ time
	\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
	Algorithmic issues for $321$-avoiding patterns containment has been investigated in
	\cite{Guillemot:Vialette:ISAAC:2009}.
	The pattern containment problem is also solvable in
	polynomial-time for separable patterns \cite{Ibarra:1997,Bose:Buss:Lubiw:1998}
	(see also \cite{Bouvel:Rossin:Vialette:CPM:2007} for LCS-like issues
	of separable permutations).
	Separable permutations are those permutations that contain neither
	$2413$ nor $3142$, and they are enumerated by the Schröder numbers
	(sequence A006318 in OEIS).
	To the best of our knowledge,
	separable permutations first arose in the work of
	Avis and Newborn~\cite{Avis:Newborn:1981},
	who showed that they are precisely the permutations which can be sorted by an
	arbitrary number of pop-stacks in series,
	where a pop-stack is a restricted form of stack in which any pop operation
	pops all items at once.
	(Notice that the separable permutations include as a special case the
	stack-sortable permutations, which avoid the pattern $231$.)

@@

A \emph{permutation} of length $n$ is a one-to-one function from an
$n$-element set to itself.
We write permutations as words 
$\sigma = \sigma_1\sigma_2\,\ldots\,\sigma_n$, whose letters are distinct
and uasually consist of the integers $12\,\ldots\,n$.
We let 
$\pi[i]$ stands for $\pi_i$.
For the sake of convenience, we let
$\pi[i:j]$ stand for 
$\pi_i\pi_{i+1}\,\ldots\,\pi_j$,
$\pi[:j]$ stand for $\pi[1:j]$ and
$\pi[i:]$ stand for $\pi[i:n]$.
As usual, we let $S_n$ denote the set of all permutations of length $n$.

@@

A \emph{bivincular pattern} (abbreviated BVP) $\sigma$
of length $k$ is a permutation in $S_k$ written in
two-line notation 
(that is the top row is $12\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\sigma_2\,\ldots\,\sigma_k$).
We have the following conditions on the top and bottom rows
of $\sigma$:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the letters corresponding to 
	$\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j$ in an occurrence of
	$\sigma$ in a permutation must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive letters.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must begin with the leftmost 
	letter of $\pi$,
	and 
	if the bottom row of $\sigma$ begins with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must end with the rightmost 
	letter of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the letters corresponding to
	$\sigma_i, \sigma_{i+1}, \ldots, \sigma_j$ in an
	occurrence of $\sigma$ in a permutation must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined 
	letters.
	Moreover, if the top row of $\sigma$ begins with
	$^\ulcorner{1}$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must begin with
	the smallest letter of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must end with
	the largest letter of $\pi$.
\end{itemize}
Given a motif $\pmotif$ we represent the elements underlined by the set $\x$. An index $i$, $1 \leq i < |\sigma|$, of an element is in $\x$ if and only if $i$ and $i+1$ need to be adjacent ($\pmotif[i]$ and $\pmotif[i+1]$ are underlined). We also allow $0$ and $|\sigma|$ to be in the set $\x$ such that if $0 \in \x$ then we have $_\llcorner{\sigma_1}$ and if $|\sigma| \in \x$ then ${\sigma_{|\sigma|}}_\lrcorner$.
We represent the index that need to be overlined by the set $\y$. An index $i$ is in $\y$ if and only if $i$ and $i+1$ need to be adjacent ($i$ and $i+1$ are overlined). We also allow $0$ and $|\sigma|$ to be in the set $\y$ such that if $0 \in \y$ then we have $^\ulcorner{1}$ and if $|\sigma| \in \y$ then ${|\sigma|}^\urcorner$.
% A bivincular pattern avoids a set of permutations $P$ if its bottom
% row $\sigma_1\sigma_2\,\ldots\,\sigma$ avoids every permutation in $P$.

@@

This section focus on the pattern matching problem when only the motif avoid $(213,231)$.

\begin{proposition}
Let $\pmotif$, $\ptext$ 
be two permutation,
with  $\pmotif$ a permutation avoiding (231,213), 
we can decide in $O(|\pmotif||\ptext|^2)$ time
and $O(|\pmotif||\ptext|^2)$ space
if $\pmotif$ 
is a motif of $\ptext$.
\end{proposition}

%			\textbf{Proof}
%			The main point is that
%			if we match a ascnt (descent) element $i_{\pmotif}$ of the motif
%			with an element $i_{\ptext}$ of the text,
%			the optimal way of matching
%			the first previous ascnt (descent)
%			that appear before an descent (ascent)
%			is to choose the
%			maximal (minimal) element possible.
%			Indeed if we do not take
%			the maximal (minimal) element,
%			we may prevent some element
%			in the future matching.\\

For the proof and the algorithm we introduce a decomposition : we decompose the stripe of the motif into factor of same letter, and label those factor from right to left. We also need to add the last index into the first factor. For each factor, we denote the left most element by $\firsta$. For example, given the motif $981237654$, $\factor{1} =7654$, $\first{1}=7$, $\factor{2}=123$, $\first{2}=1$, $\factor{3}=98$ and $\first{9}$. \\

\begin{proof}
Given an ascent (descent) factor $\factor{i}$ of a motif $\pmotif$, if we want to match the suffix of $\pmotif$ starting by $\factor{i}$, $\factor{i}\factor{i-1}\cdots\factor{1}$ and given that we know every match of $\factor{i-1}\cdots\factor{1}$, the optimal match is to choose the match of $\factor{i-1}\cdots\factor{1}$ that minimize (maximize) the maximal (minimal) element of the match of $\factor{i-1}\cdots\factor{1}$ and that allow a match of factor $\factor{i}$. In other word if we want to match $\pmotif[\first{i}:]$ with $\ptext[j:]$ such that $\first{i}$ is matched with element at index $j$ in $\ptext$, the optimal way of matching $\pmotif[\first{i-1}:]$ is to choose the minimal (maximal) element at index $j'$ of $\ptext[j:]$ such that : (1) $\pmotif[\first{i-1}:]$ is matched with $\ptext[j':]$ and element $\first{i-1}$ is matched with element at index $j'$, (2) the factor $\factor{i}$ is matched with $\ptext[j,j'-1]$ and element $\first{i}$ is matched with element at index $j$, (3) Every elements of the match of (2) if inferior (superior) to the minimal (maximal) element of the match of (1). Indeed when we will match $\factor{i+1}\factor{i}\factor{i-1}\cdots\factor{1}$, every element of $\factor{i+1}$ must be matched to element superior (inferior) to every element matched element of $\factor{i}\factor{i-1}\cdots\factor{1}$, particularly its must be superior (inferior) to the maximal (minimal) element of the match of $\factor{i}\factor{i-1}\cdots\factor{1}$.

@@

There exists a match of $\pmotif$ in $\ptext$ if and only if
there exists a $LM(n,i)\neq 0$ for $1 \leq i \leq |\ptext|$,
with $n$ the number of factor in $\pmotif$.
Moreover the basic case can be computed in $O(|\ptext|^2)$
and the induction on $O(|\pmotif||\ptext|^2)$.
\end{proof}

@@

\section{Pattern Matching With Bivincular Motif Avoiding (231,213)}	
\label{section:bivincular}

This section gives an algorithm to detect bivincular permutation in polynomial time in the case where the motif avoid $(213,231)$.
We denote $\pbmotif=\bmotif$ the bivincular motif.

\begin{proposition}
Let $\pbmotif$ be a bivincular motif such that $\pmotif$
avoid (231,213), 
$\ptext$ a permutation,
we can decide in $O(|\pmotif||\ptext|^3)$ time
and $O(|\ptext|^3)$ space
if $\pbmotif$
appear in $\ptext$.
\end{proposition}

@@

$
\PM{\pbmotif}{\ptext}{\min}{\max}{i}{j}=
\begin{cases}
	true 	& \text{If $\pbmotif[i:]$ is a motif of $\ptext[j:]$}\\
			& \text{with every element in}\\ 
			&[\min,\max]\\

@@

There are 6 cases to consider to search a match depending of if $0 \in \x$, $0 \in \y$ and if the first element of the motif is ascent or descent :
\begin{itemize}
	\item If $0 \in \x$ and $0 \notin \y$ there exists a match if and only if there is a match starting with left most element of the text:\\ $\bigcup_{0<\min<\max<n_\ptext}
	\PM{\pbmotif}{\ptext}{\min}{\max}{0}{0}$ is true.

	\item If $n_\pmotif \notin \x$ and $0 \notin \y$ there exists a match if and only if there exists a match stating at any index: \\	$\bigcup_{0<k<n_\ptext,0<\min<\max<n_\ptext}\PM{\pbmotif}{\ptext}{\min}{\max}{0}{k}$ is true.

	\item If $0 \in \x$, $0 \in \y$ and the first element of the motif is ascent there exists a match if and only if there is a match starting with the left most element  of $\ptext$, and the left most element  of $\ptext$ is the minimal element  of $\ptext$:\\ $\bigcup_{0<\min<\max<n_\ptext}\PM{\pbmotif}{\ptext}{\min}{\max}{0}{0}$  is true and the left most element of $\ptext$ is the minimal element of $\ptext$.

	\item If $n_\pmotif \notin \x$, $0 \in \y$ and the first element of the motif is ascent there exists a match if and only if there exists a match stating at the index of the minimal element of $\ptext$ : if $j_{\min}$ is the index of the minimal element then \\	$\bigcup_{0<\min<\max<n_\ptext}\PM{\pbmotif}{\ptext}{\min}{\max}{0}{j_{\min}}$ is true.

	\item If $0 \in \x$, $0 \notin \y$  and the first element of the motif is descent then there exists a match if and only if there is a match starting with the left most element  of $\ptext$ and the left most element  of $\ptext$ is the maximal element  of $\ptext$:\\ $\bigcup_{0<\min<\max<n_\ptext}\PM{\pbmotif}{\ptext}{\min}{\max}{0}{0}$ is true and the left most element of $\ptext$ is the maximal element of $\ptext$.

	\item If $n_\pmotif \notin \x$, $0 \notin \y$ and the first element of the motif is descent there exists a match if and only if there exists a match stating at the index of the maximal element of $\ptext$ : if $j_{\max}$ is the index of the maximal element then  \\	$\bigcup_{0<\min<\max<n_\ptext}\PM{\pbmotif}{\ptext}{\min}{\max}{0}{j_{\max}}$ is true.

\end{itemize}

\end{proof}
